Acid-induced protein gels: from gelation to stress-induced failure

### 2016-06-08 seminaire_PIMM.tex
Adapt to iLM template

\documentclass[xcolor=table]{beamer}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage[super]{nth}
%\usetheme{Boadilla}
%\usecolortheme{rose}
%\usecolortheme{crane}
\usefonttheme{structuresmallcapsserif}
\setbeamertemplate{navigation symbols}{}

@@

\usecolortheme{rose}
%\useinnertheme[shadow]{circles}
\usecolortheme{whale}
%\useoutertheme{infolines}

\usecolortheme[named=Accent1]{structure}

\setbeamercolor{alerted text}{fg=Accent2}
%\setbeamercolor{palette primary}{fg=white}
%\setbeamercolor{palette secondary}{bg=Accent1}
%\setbeamercolor{palette tertiary}{bg=Accent2,fg=white}

\setbeamerfont{page number in head/foot}{size=\large}
\setbeamercolor{page number in head/foot}{fg=Main}
% page/total
%\setbeamertemplate{footline}[frame number]
% pas de total
\setbeamertemplate{footline}{%
    	\hfill%
	\usebeamercolor[fg]{page number in head/foot}%
	\usebeamerfont{page number in head/foot}%
	\insertframenumber\kern1em\vskip2pt%
}

\setbeamersize{text margin left=1em}
\setbeamersize{text margin right=1em}

%font
\usepackage[T1]{fontenc}
\usepackage[oldstylenums]{kpfonts}

@@

\tikzset{every picture/.style={beamer externalizing}}
\tikzexternalize
\tikzsetexternalprefix{fig_presentation/}
%\tikzset{external/optimize=false}
%\tikzset{external/force remake}

@@

\institute[iLM]{Institut Lumière Matière, Lyon, France}
\title[Yoghurt under stress]{Acid-induced protein gels: from gelation to stress-induced failure}
\author[M. Leocmach]{Mathieu Leocmach}
\date{10 June 2015}
\titlegraphic{
	\begin{tabu}{X[c]X[c]X[c]X[c]}
		\includegraphics[height=3\baselineskip]{Chris}&
		\includegraphics[height=3\baselineskip]{Thomas}&
		\includegraphics[height=3\baselineskip]{Thibaut}&
		\includegraphics[height=3\baselineskip]{Seb}\\
		Christophe Perge & Thomas\linebreak Gibaud & Thibaut\linebreak Divoux & Sebastien Manneville\\
	\end{tabu}

@@

\begin{frame}[plain]
	\titlepage
\end{frame}

@@

\begin{frame}{Texture as important as taste}
\tikzsetnextfilename{texture}%
\begin{tikzpicture}[inner sep=0, callout pointer segments=3]
\node(ishichan) {\includegraphics[height=8\baselineskip]{ishichan}};
\node[above=of ishichan] {\includegraphics[height=8\baselineskip]{calamar}};
\node[cloud callout,anchor=west, draw, callout relative pointer={(-30:2\baselineskip)}] at ($(ishichan.north)+(-0.5\textwidth,0)$) {\includegraphics[height=5\baselineskip]{aldente}};
\node[cloud callout, aspect=0.75, anchor=east, draw, callout relative pointer={(-150:2\baselineskip)}] at ($(ishichan.north)+(0.5\textwidth-10pt,0)$) {\includegraphics[height=5\baselineskip]{chewie}};
\end{tikzpicture}
\end{frame}

@@

\column{0.475\textwidth}
\begin{block}{Brittle fracture}
\includegraphics[height=6\baselineskip]{spaghetti_breaking}\hfill
\includegraphics[height=6\baselineskip]{Tea_bowl_fixed_in_the_Kintsugi_method}
\begin{itemize}
\item concrete
\item asphalt
\item glass
\end{itemize}

@@

%\tikzset{external/force remake}
\begin{frame}{The simplest yoghurt}
\begin{block}{Acid-set protein gel}
\begin{itemize}
\item Water (\SI{30}{\celsius})
\item Sodium caseinate (milk protein) 4\%
\raisebox{0.6\normalbaselineskip}[0pt][0pt]{$\left.\rule{0pt}{1.1\normalbaselineskip}\right(\}$ stable solution}
\item Glucono-$\delta$-lactone (GDL) 1\% $\Rightarrow$ slow homogeneous acidification
\end{itemize}
\end{block}

@@

\begin{frame}{Linear rheology: Power law (visco)elastic solid}
\[\text{stress}\rightarrow \sigma = G \gamma\leftarrow\text{strain}\]
\[\textcolor{Accent1}{\text{storage}\rightarrow} G^\prime + \imath G^{\prime\prime} \textcolor{Accent2}{\leftarrow\text{loss}}\]
\tikzsetnextfilename{powerlaw}
\begin{tikzpicture}
\begin{loglogaxis}[
	height=0.8\textheight,
	width=\textwidth,
	xlabel={Frequency (\si{\hertz})}, ylabel={\textcolor{Accent1}{Storage} and \textcolor{Accent2}{loss} moduli (\si{\pascal})},
	domain={6e-2:70},
	]
	\addplot[Accent1, only marks, mark=*] table{freqsweep_Y265_cas4_GDL1.txt};
	\addplot[Accent2, only marks, mark=o] table[y=LossModulus]{freqsweep_Y265_cas4_GDL1.txt};
	\addplot[Main, no marks]{300*x^0.15}  node[midway, below right] {$G^\prime\sim G^{\prime\prime}\sim f^{0.15}$};
%	\addplot+[only marks, mark=*, forget plot] table{freqsweep_Y265_cas4_GDL1.txt} node[anchor=base east] {1\%};
%	\addplot+[only marks, mark=o, forget plot] table[y=LossModulus]{freqsweep_Y265_cas4_GDL1.txt} node[anchor=base east] {1\%};
%	\addplot{590*x^0.14};
%	\addplot+[only marks, mark=*, forget plot] table{freqsweep_Y277_cas4_GDL1.25.txt} node[anchor=base east] {1.25\%};
%	\addplot+[only marks, mark=o, forget plot] table[y=LossModulus]{freqsweep_Y277_cas4_GDL1.25.txt} node[anchor=base east] {1.25\%};
%	\addplot{350*x^0.12};
%	\addplot+[only marks, mark=*, forget plot] table{freqsweep_Y275_cas4_GDL1.5.txt} node[anchor=base east] {1.5\%};
%	\addplot+[only marks, mark=o, forget plot] table[y=LossModulus]{freqsweep_Y275_cas4_GDL1.5.txt} node[anchor=base east] {1.5\%};
%	\addplot{246*x^0.1};
%	\addplot+[only marks, mark=*, forget plot] table{freqsweep_Y268_cas4_GDL2.txt} node[anchor=base east] {2\%};
%	\addplot+[only marks, mark=o, forget plot] table[y=LossModulus]{freqsweep_Y268_cas4_GDL2.txt} node[anchor=base east] {2\%};
%	\addplot{163*x^0.067};
%	\addplot+[only marks, mark=*, forget plot] table{freqsweep_Y270_cas4_GDL3.txt} node[anchor=base east] {3\%};
%	\addplot+[only marks, mark=o, forget plot] table[y=LossModulus]{freqsweep_Y270_cas4_GDL3.txt} node[anchor=base east] {3\%};
%	\addplot{109*x^0.044};
%	\addplot+[only marks, mark=*, forget plot] table{freqsweep_Y271_cas4_GDL4.txt} node[anchor=base east] {4\%};
%	\addplot+[only marks, mark=o, forget plot] table[y=LossModulus]{freqsweep_Y271_cas4_GDL4.txt} node[anchor=base east] {4\%};
%	\addplot{76*x^0.035};

\end{loglogaxis}
\end{tikzpicture}
\end{frame}

@@

\begin{frame}{Can we generalise to biogels? \textup{\normalsize(physical gels
$E_{bond}\gg k_B T$)}}
\begin{tabu}{cX[c]c}
\includegraphics[height=0.3\textheight]{cell_mech} &
\includegraphics[height=0.3\textheight]{electrophoresis} &
\includegraphics[height=0.3\textheight]{cosmetics} \\
Cell mechanics & Electrophoresis & Cosmetics\\
\end{tabu}
\begin{tabu}{X[c]X[c]}
\includegraphics[height=0.3\textheight]{bacterial_culture} &
\includegraphics[height=0.3\textheight]{food} \\
Bacterial culture & Food\\
\end{tabu}

@@

\begin{frame}{Acidifying casein solutions with \textup{HCl}}
\tikzsetnextfilename{dosage}
\input{fig_presentation/dosage.pgf}
\end{frame}

\begin{frame}{Over-acidified protein gels: beyond isoelectric pH}
	\tikzsetnextfilename{prise_cas4}
	\input{fig_soundbite_palavas/prise_cas4.pgf}
	\begin{itemize}
	\item[$\Rightarrow$] still solid
	\item From critical to super-critical gels?
	\end{itemize}
\end{frame}

\begin{frame}{Linear rheology, from power-law to soft glassy}
	\tikzsetnextfilename{sweep_cas4}
	\input{fig_soundbite_palavas/sweep_cas4.pgf}
	\begin{itemize}
	\item Near isoelectric, power law rheology $G^\prime, G^{\prime\prime}$ typical of critical gels 
	\item Over-acidification lowers exponent and creates a minimum in $G^{\prime\prime}$
	\item Transition to soft-glassy behaviour \\\textit{\footnotesize Sollich, PRE (1998); Mason, Curr. Opin. Colloid Interface Sci. (1999)}
	\end{itemize}
\end{frame}

@@

\begin{frame}{Microscructure: fluorescent confocal microscopy}
	\tikzsetnextfilename{structure_confocal}
	\input{fig_soundbite_palavas/structure_confocal.pgf}
	\vspace{-1em}
	\begin{itemize}
	\item Maximum $G^\prime$ corresponds to maximum contrast between phases
	\item Pore size $\xi$ decrease during over-acidification
	\item<2> Casein aggregates swell $\sim$ over compressed gel $\Rightarrow$ soft glassy
	\end{itemize}
\end{frame}

\begin{frame}{Three creep regimes + softness\hfill\textsc{gdl} 4\%}
\vfill
	\tikzsetnextfilename{creep_cas4_gdl4}
	\input{fig_presentation/creep_cas4_gdl4.pgf}
\begin{itemize}
\item reversible, homogeneous creep $\rightarrow$ irreversible fracture growth
\item nucleation regime \alert{NOT universal}
\begin{itemize}
	\item possible to reform bond?
	\item friction between swollen strands?
	\item glassy dissipation?
\end{itemize}
\end{itemize}

@@

\begin{frame}{Basquin law rescaling}
	\tikzsetnextfilename{basquin_rescale}
	\input{fig_soundbite_palavas/basquin.pgf}
\end{frame}

\begin{frame}{Monkman-Grant: Predict failure from nucleation}
	\tikzsetnextfilename{MonkmanGrant}
	\input{fig_soundbite_palavas/MonkmanGrant.pgf}
\end{frame}

\begin{frame}{Wavelength}
	\tikzsetnextfilename{lambda}
	\input{fig_soundbite_palavas/lambda.pgf}
	\textcolor{Accent1}{$\blacktriangleright$} $\lambda\propto e$ compatible with nucleation rate

>>>>

\documentclass[xcolor=table]{beamer}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage[super]{nth}
%\usetheme{Boadilla}
%\usecolortheme{rose}
%\usecolortheme{crane}
%\usefonttheme{structuresmallcapsserif}
\setbeamertemplate{navigation symbols}{}

@@

\mode<presentation>{\usetheme{ilm}}

%\usecolortheme{rose}
%%\useinnertheme[shadow]{circles}
%\usecolortheme{whale}
%%\useoutertheme{infolines}
%
%\usecolortheme[named=Accent1]{structure}
%
%\setbeamercolor{alerted text}{fg=Accent2}
%\setbeamercolor{palette primary}{fg=white}
\setbeamercolor{palette secondary}{bg=ilmcolor!10!white}
%\setbeamercolor{palette tertiary}{bg=Accent2,fg=white}

%\setbeamerfont{page number in head/foot}{size=\large}
%\setbeamercolor{page number in head/foot}{fg=Main}
%% page/total
%%\setbeamertemplate{footline}[frame number]
%% pas de total
%\setbeamertemplate{footline}{%
%    	\hfill%
%	\usebeamercolor[fg]{page number in head/foot}%
%	\usebeamerfont{page number in head/foot}%
%	\insertframenumber\kern1em\vskip2pt%
%}
%
%\setbeamersize{text margin left=1em}
%\setbeamersize{text margin right=1em}

%font
\usepackage[T1]{fontenc}
\usepackage{times}

@@

\tikzset{every picture/.style={beamer externalizing}}
\tikzexternalize
\tikzsetexternalprefix{fig_iLM/}
%\tikzset{external/optimize=false}
%\tikzset{external/force remake}

@@

\institute[Lyon]{iLM, Univ. Claude Bernard Lyon 1 and CNRS, France}
\title[Yoghurt under stress]{Acid-induced protein gels: from gelation to stress-induced failure}
\author[M. Leocmach]{Mathieu Leocmach}
\date{9 June 2016}
\titlegraphic{
	\begin{tabu}{X[c]X[c]X[c]X[c]}
		\includegraphics[height=3\baselineskip]{Chris}&
		\includegraphics[height=3\baselineskip]{Thomas}&
		\includegraphics[height=3\baselineskip]{Thibaut}&
		\includegraphics[height=3\baselineskip]{Seb}\\
		Christophe Perge & Thomas\linebreak Gibaud & Thibaut\linebreak Divoux & Sebastien Manneville\\
	\end{tabu}

@@

\begin{frame}{\pgfuseimage{cnrs-logo}\hspace*{0.3cm}\pgfuseimage{ucbl-logo}}%[plain]
	\titlepage
\end{frame}

@@

\begin{frame}{Texture as important as taste}
\vspace{1.25\baselineskip}
\tikzsetnextfilename{texture}%
\begin{tikzpicture}[inner sep=0, callout pointer segments=3]
\node(ishichan) {\includegraphics[height=8\baselineskip]{ishichan}};
\node[above=1em of ishichan] {\includegraphics[height=8\baselineskip]{calamar}};
\node[cloud callout,anchor=west, draw, callout relative pointer={(-30:2\baselineskip)}] at ($(ishichan.north)+(-0.5\textwidth-10pt,0)$) {\includegraphics[height=5\baselineskip]{aldente}};
\node[cloud callout, aspect=0.75, anchor=east, draw, callout relative pointer={(-150:2\baselineskip)}] at ($(ishichan.north)+(0.5\textwidth-10pt,0)$) {\includegraphics[height=5\baselineskip]{chewie}};
\end{tikzpicture}
\end{frame}

@@

\column{0.5\textwidth}
\begin{block}{Brittle fracture}
\includegraphics[height=6\baselineskip]{spaghetti_breaking}\hfill
\includegraphics[height=6\baselineskip]{Tea_bowl_fixed_in_the_Kintsugi_method}
\begin{itemize}
\item concrete
\item asphalt
\item glass
\end{itemize}

@@

%\tikzset{external/force remake}
\begin{frame}{The simplest yoghurt}
\begin{block}{Acid-set protein gel}
\begin{itemize}
\item Water (\SI{30}{\celsius})
\item Sodium caseinate (milk protein) 4\%
\raisebox{0.6\normalbaselineskip}[0pt][0pt]{$\left.\rule{0pt}{1.1\normalbaselineskip}\right\}$ stable solution}
\item Glucono-$\delta$-lactone (GDL) 1\% $\Rightarrow$ slow homogeneous acidification
\end{itemize}
\end{block}

@@

\begin{frame}{Power law linear rheology}
\[\text{stress}\rightarrow \sigma = G \gamma\leftarrow\text{strain}\]
\[\textcolor{Accent1}{\text{storage}\rightarrow} G^\prime + \imath G^{\prime\prime} \textcolor{Accent2}{\leftarrow\text{loss}}\]
\tikzsetnextfilename{powerlaw}
\begin{tikzpicture}
\begin{loglogaxis}[
	height=0.8\textheight,
	width=\textwidth,
	xlabel={Frequency (\si{\hertz})}, ylabel={\textcolor{Accent1}{Storage} and \textcolor{Accent2}{loss} moduli (\si{\pascal})},
	domain={6e-2:70},
	]
	\addplot[Accent1, only marks, mark=*] table{freqsweep_Y265_cas4_GDL1.txt};
	\addplot[Accent2, only marks, mark=o] table[y=LossModulus]{freqsweep_Y265_cas4_GDL1.txt};
	\addplot[Main, no marks]{300*x^0.15}  node[midway, below right] {$G^\prime\sim G^{\prime\prime}\sim f^{0.15}$};
	\node[below right] at (rel axis cs:0.1,0.9) {(visco)elastic solid};
\end{loglogaxis}
\end{tikzpicture}
\end{frame}

@@

\begin{frame}{Can we generalise to biogels?}
\hfill\textup{\footnotesize\usebeamercolor[fg]{titlelike}(physical gels
$E_{bond}\gg k_B T$)}
\begin{tabu}{cX[c]c}
\includegraphics[height=0.3\textheight]{cell_mech} &
\includegraphics[height=0.3\textheight]{electrophoresis} &
\includegraphics[height=0.3\textheight]{cosmetics} \\
Cell mechanics & Electrophoresis & Cosmetics\\
\end{tabu}
\begin{tabu}{X[c]X[c]}
\includegraphics[height=0.3\textheight]{bacterial_culture} &
\includegraphics[height=0.3\textheight]{food} \\
Bacterial culture & Food\\
\end{tabu}

@@

\begin{frame}{Acidifying casein solutions with \textup{HCl}}
\vspace{1.25\baselineskip}
\tikzsetnextfilename{dosage}%
\input{fig_presentation/dosage.pgf}
\end{frame}

\begin{frame}{Over-acidified protein gels}
	\hfill beyond isoelectric pH
	\tikzsetnextfilename{prise_cas4}
	\input{fig_soundbite_palavas/prise_cas4.pgf}
	\begin{itemize}
	\item[$\Rightarrow$] still solid
	\item From critical to super-critical gels?
	\end{itemize}
\end{frame}

\begin{frame}{Linear rheology $\rightarrow$ to soft glassy}
	\vspace{\baselineskip}
	\tikzsetnextfilename{sweep_cas4}%
\input{fig_soundbite_palavas/sweep_cas4.pgf}
	\begin{itemize}
	\item Near isoelectric, power law rheology typical of critical gels 
	\item Over-acidification: exponent $\searrow$ \hfill\textcolor{ilmorange}{\&}\hfill minimum in $G^{\prime\prime}$
	\item Transition from power-law to soft-glassy behaviour \\\textit{\footnotesize Sollich, PRE (1998); Mason, Curr. Opin. Colloid Interface Sci. (1999)}
	\end{itemize}
\end{frame}

@@

\begin{frame}{Microscructure: confocal microscopy}
	\vspace{\baselineskip}
	\tikzsetnextfilename{structure_confocal}%
\input{fig_soundbite_palavas/structure_confocal.pgf}
	\vspace{-1em}
	\begin{itemize}
	\item Maximum $G^\prime$\hfill\textcolor{ilmorange}{$\leftrightarrow$}\hfill Maximum contrast between phases
	\item Pore size $\xi$ decrease during over-acidification
	\item<2> Casein aggregates swell $\sim$ over compressed $\Rightarrow$ soft glassy
	\end{itemize}
\end{frame}

\begin{frame}{3 creep regimes + softness\textsc{gdl} 4\%}
\vfill
	\tikzsetnextfilename{creep_cas4_gdl4}
	\input{fig_presentation/creep_cas4_gdl4.pgf}
\begin{itemize}
\item reversible, homogeneous creep $\rightarrow$ irreversible fracture growth
\item nucleation regime \alert{NOT universal}
\begin{itemize}
	\item possible to reform bond?
	\item friction between swollen strands?
	\item glassy dissipation?
\end{itemize}
\end{itemize}

@@

\begin{frame}{Basquin law rescaling}
	\vspace{0.5\baselineskip}
	\tikzsetnextfilename{basquin_rescale}%
	\input{fig_soundbite_palavas/basquin.pgf}
\end{frame}

\begin{frame}{Monkman-Grant}
	\vspace{0.5\baselineskip}
	\hfill Predict failure from nucleation
	\tikzsetnextfilename{MonkmanGrant}
	\input{fig_soundbite_palavas/MonkmanGrant.pgf}
\end{frame}

\begin{frame}{Wavelength}
	\vspace{1.25\baselineskip}
	\tikzsetnextfilename{lambda}%
	\input{fig_soundbite_palavas/lambda.pgf}
	\textcolor{Accent1}{$\blacktriangleright$} $\lambda\propto e$ compatible with nucleation rate
